Estimating demographic bias on tests of children's early vocabulary

### 2022-05-05 paper/cogsci_final.Rmd
CogSci revisions

author-information: > 
    \author{{\large \bf George Kachergis (kachergis@stanford.edu)}
    \AND {\large \bf Nathan Francis (nathan99@stanford.edu)}
    \AND {\large \bf Michael C. Frank (mcfrank@stanford.edu)} \\ Department of Psychology, Stanford Unviersity \\ Stanford, CA 94305 USA
    }

abstract: >
    Children's early language skill has been linked to later educational outcomes, making
    it important to measure early language accurately. Parent-reported instruments such as the 
    Communicative Development Inventories (CDIs) have been shown to provide reliable and valid
    measures of children's aggregate early language skill. However, CDIs contain
    hundreds of vocabulary items, some of which may not be heard (and thus learned) equally
    often by children of varying backgrounds. This study used a database of American English 
    CDIs to identify words demonstrating strong bias for particular demographic groups of children, on 
    dimensions of sex (male vs. female), race (white vs. non-white), and maternal education 
    (high vs. low). For each dimension, we identified dozens of strongly biased items, and showed that 
    eliminating these items reduced the expected ability difference between groups. Additionally,
    we investigated how well the relative frequency of words spoken to young girls vs. boys predicted 
    sex-based word learning bias, and discuss possible sources of demographic differences in early word learning.

@@

output: word_document
#final-submission: \cogscifinalcopy
---

@@

CDIs are a useful tool for measuring demographic effects because of their wide use across languages. 
But their use in this context poses a chicken and egg problem. 
While it is possible that there is an underlying difference in mean language ability between demographic groups, CDIs could also have bias for or against particular groups based on the words they include.

@@

Demographic differences in language ability are likely to be at least partially explained by differences in linguistic input received by children in different groups.
Indeed, input quantity (total daily tokens) and some measures of quality (e.g., lexical diversity: ratio of word types vs. tokens) have often been predictive of language learning outcomes in  demographic studies [@rowe2009differences; @huttenlocher1991early].
We used a corpus-based analysis to investigate the extent to which word frequency in child-directed speech to male vs. female children was predictive of the amount of sex-based bias shown by CDI items. 
Similar to the approach taken by @braginsky2016gender, we used the CHILDES corpus of transcripts from dyadic play sessions [@macwhinney2000childes], which are labeled with the sex of the target child, but not other demographic variables.
We found a total of 5213750 female-directed tokens and 6091950 male-directed tokens that matched 662 of the 680 CDI:WS items, normalized the word frequencies to tokens per million for each target sex, and calculated the log-odds of each word being spoken to a female (i.e., $log(p_{f} / p_{m}$), meaning unbiased words will have log-odds of 0, while those spoken more often to females will have log-odds $>0$, and those spoken more often males will have log-odds $<0$.
For example "doll" was spoken 156 times to females, and 138 times to males, thus $p_{f} = 156/(156+138) = .53$, $p_{m} = 1-p_{f} = 0.47$, and $log(p_{f}/p_{m})$ = 0.12), while "police" was spoken 138 times to females, and 225 times to males, and thus has log-odds = -0.49).
Overall, the correlation between the log-odds of a word being spoken to a female vs. a male child and the size of the female (vs. male) advantage for that CDI word was modest, but significant ($r=0.18$, $p<.001$). 
Some of the sex bias seen in CDI items is related to differences in language input received by girls vs. boys -- though of course this differential input could be elicited due to different interests.

@@

This investigation is only a first step in measuring demographic bias in the items on the CDI:WS.
More research is needed to determine whether CDI items that show DIF are more related to environmental frequencies, as we found only a modest association between the relative frequency of words in child-directed speech to boys vs. girls and the amount of sex-based DIF.
Examining the relative frequency of CDI words across households with varying maternal education and race may reveal similarly weak associations, or there may be strong associations only among particular categories of words.
It may be that particular activity contexts--popular with some demographic groups, and not others--may be predictive. 
For example, many of the extrema favoring high-ME children are animals and animal sounds (e.g., "quack quack", "woof woof", "baa baa", "duck", "sheep", "giraffe", "zebra"): do high-SES households visit the zoo more often, or perhaps engage in other activities related to naming animals and noises they make (e.g., animal noise toys)?
If certain activities are driving early word learning for high-ME children, what are the activities (and associated vocabulary) that low-SES households are instead engaging in?
Critically this enterprise will require a greater diversity of publicly-available corpora of child language; we view the lack of such data as a substantial issue.

@@

[Redacted for anonymous review.]
<!--We thank members of the Language and Cognition lab for their feedback.
This research was supported by the Stanford Maternal and Child Health Research Institute.  
-->

>>>>

author-information: > 
    \author{{\large \bf George Kachergis, Nathan Francis, and Michael C. Frank}
    \AND {\large \tt  \{kachergis, nathan99, mcfrank\} @stanford.edu} \\ Department of Psychology, Stanford Unviersity \\ Stanford, CA 94305 USA
    }

abstract: >
    Children's early language skill has been linked to later educational outcomes, making
    it important to measure early language accurately. Parent-reported instruments such as the 
    Communicative Development Inventories (CDIs) have been shown to provide reliable and valid
    measures of children's aggregate early language skill. However, CDIs contain
    hundreds of vocabulary items, some of which may not be heard (and thus learned) equally
    often by children of varying backgrounds. This study used a database of American English 
    CDIs to identify words demonstrating strong bias for particular demographic groups of children, on 
    dimensions of sex (male vs. female), race (white vs. non-white), and maternal education 
    (high vs. low). For each dimension, we identified dozens of strongly biased items, and showed that 
    eliminating even some of these items can reduce the magnitude of differences between groups. 
    Additionally, we investigated how well the relative frequency of words spoken to young girls vs.
    boys predicted sex-based word learning bias, and discuss possible sources of demographic
    differences in early word learning.

@@

output: cogsci2016::cogsci_paper
final-submission: \cogscifinalcopy
---

@@

CDIs are a useful tool for measuring demographic effects because of their ease of use and wide distribution to different groups. 
But their use in this context poses a chicken and egg problem. 
While it is possible that there is an underlying difference in mean language ability (or vocabulary size) between demographic groups, CDIs could also have bias for or against particular groups based on the particular words they include, which different groups may have a varying amount of exposure to.

@@

Demographic differences in language ability are likely to be at least partially explained by differences in linguistic input received by children in different groups.
Indeed, input quantity (total daily tokens) and some measures of quality (e.g., lexical diversity: ratio of word types vs. tokens) have often been predictive of language learning outcomes in  demographic studies [@rowe2009differences; @huttenlocher1991early].
We used a corpus-based analysis to investigate the extent to which word frequency in child-directed speech to male vs. female children was predictive of the amount of sex-based bias shown by CDI items. 
Similar to the approach taken by @braginsky2016gender, we used the CHILDES corpus of transcripts from dyadic play sessions [@macwhinney2000childes], which are labeled with the sex of the target child, but not other demographic variables.
We found a total of 5213750 female-directed tokens and 6091950 male-directed tokens that matched 662 of the 680 CDI:WS items, normalized the word frequencies to tokens per million for each target sex, and calculated the log-odds of each word being spoken to a female (i.e., $log(p_{f} / p_{m}$), meaning unbiased words will have log-odds of 0, while those spoken more often to females will have log-odds $>0$, and those spoken more often males will have log-odds $<0$.
For example "doll" was spoken 156 times to females, and 138 times to males, thus $p_{f} = 156/(156+138) = .53$, $p_{m} = 1-p_{f} = 0.47$, and $log(p_{f}/p_{m})$ = 0.12), while "police" was spoken 138 times to females, and 225 times to males, and thus has log-odds = -0.49).
Overall, the correlation between the log-odds of a word being spoken to a female vs. a male child and the size of the female (vs. male) advantage for that CDI word was weak, but significant ($r=0.18$, $p<.001$). 
Some of the sex bias seen in CDI items is related to differences in language input received by girls vs. boys -- though of course this differential input could be elicited due to different interests.

@@

This investigation is only a first step in measuring demographic bias in the items on the CDI:WS.
More research is needed to determine whether CDI items that show DIF are more related to environmental frequencies, as we found a weak though reliable association between the relative frequency of words in child-directed speech to boys vs. girls and the amount of sex-based DIF.
Examining the relative frequency of CDI words across households with varying maternal education and race may reveal similarly weak associations, or there may be strong associations only among particular categories of words.
It may be that particular activity contexts--popular with some demographic groups, and not others--may be predictive. 
For example, many of the extrema favoring high-ME children are animals and animal sounds (e.g., "quack quack", "woof woof", "baa baa", "duck", "sheep", "giraffe", "zebra"): do high-SES households visit the zoo more often, or perhaps engage in other activities related to naming animals and noises they make (e.g., animal noise toys)?
If certain activities are driving early word learning for high-ME children, what are the activities (and associated vocabulary) that low-SES households are instead engaging in?
Critically this enterprise will require a greater diversity of publicly-available corpora of child language; we view the lack of such data as a substantial issue.

@@

We thank members of the Language and Cognition lab and three anonymous reviewers for their feedback.
